---
title: "Pathway Analysis w/ MSigDB"
author: "D. Ford Hannum"
date: "`r Sys.Date()`"
output: 
        html_document: 
          toc: yes
          toc_float: yes
          theme: united
          highlight: tango
          # code_folding: hide
          number_sections: true
          keep_md: no
          toc_depth: 5
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# Set the directory to where everything is located
knitr::opts_knit$set(root.dir = '/Users/dfhannum/Desktop/OphthalmApp/') # sets it for all code chunks

library(Seurat) # for scRNA functions and plots
library(ggplot2) # for custom plotting
library(data.table) # data file processing and tstrsplit
library(clusterProfiler) # pathway analysis

library(org.Mm.eg.db) # <org.Hs.eg.db> human or <org.Rn.eg.db> rat

library(msigdbr) # access to MSigDB
```

# Introduction

In this document we are going to use an already created Seurat scRNA object. We are then going to identify differentially expressed genes between conditions within clusters, and submit those markers for over-representation (enrichment) analysis with MSigDB. We will then make some plots representing the most significant pathways.


# Data

I'm using a previously published dataset that I've reanalyzed for other example scripts.

```{r}
# Loading in the data
obj <- readRDS('./data/seurat_obj_full.Rds') 
```

```{r}
# Visualizing the data
DimPlot(obj)
```

We are going to rerun the clustering to identify a smaller number of clusters. Many of these clusters will share similar transcriptomic profiles. For the speed/visualizations of my downstream example it'll be easier working with fewer clusters.

```{r}
obj <- FindClusters(obj, resolution = .05)
```

```{r}
# Projecting the data in umap space
DimPlot(obj)
```

We can also look at the umap for mutant and WT separately. 

```{r}
DimPlot(obj, split.by = 'orig.ident')
```

# Finding Condition DEGs within Clusters

We will calculate differentially expressed genes between conditions for each cluster.

We can create a new variable that combines our condition (orig.ident for this data) and cluster.

```{r}
table(Idents(obj), 
      obj$orig.ident)

obj$cond_by_cluster <- paste0(obj$orig.ident,'_', Idents(obj))
```


```{r}
# Creating empty list
results <- list()

# Looping through clusters
for (cluster in levels(Idents(obj))){
  # print(cluster)
  
  markers <- FindMarkers(object = obj,
                         test.use = 'wilcox',
                         group.by = 'cond_by_cluster', # variable to group by
                         ident.1 = paste0('WT_',cluster), # wt in cluster
                         ident.2 = paste0('MUT_',cluster), # mut in cluster
                         logfc.threshold = .2) # a fc threshold to limit to sig. results and to run faster
  
  # Adding it to the results list
  results[[paste0('c', cluster)]] <- markers
}
```


## Alternative

If you have more than one condition the above code will not work. Here you'll need to decide if you're running one genotype vs all others (ex. geno1 vs geno2 + geno3) or in a pairwise manner (ex. geno1 vs geno2, geno1 vs geno3). 

For the one vs all others the below code should work instead of the previous code chunk.

```{r}

# Creating a new random variable with three genotypes by cluster
obj$geno <- sample(c('geno1','geno2','geno3'), dim(obj)[2], replace = T)

results2 <- list()

# demo'ing through one cluster
for (i in levels(Idents(obj))){
  print(i)
  for (geno in unique(obj$geno)){
    # This pulls compares geno1 to geno2 & geno3 within cluster i
    obj$temp_label <- ifelse(obj$seurat_clusters != i, 'other',
                             ifelse(obj$geno == geno, geno,
                                    'other_geno'))
    
    markers <- FindMarkers(obj,
                           group.by = 'temp_label',
                           ident.1 = geno,
                           ident.2 = 'other_geno',
                           logfc.threshold = .2
                           # ... other params you want to include
                           )
    
    results2[[paste0('c',i,'_',geno)]] <- markers
    
  }
  
  break # breaking out of the for loop
}

```


## Extracting DEGs

Next we are going to extract the differentially expressed genes for each cluster

```{r}
# Creating an empty list to add to
marker_list <- list()

# A for loop to loop through the clusters
# You can replace obj$seurat_clusters with your cluster identity
for (i in names(results)){
  
  # Print statements are always useful in for loops to track progress and to 
  # make sure it's doing what it's suppose to
  # print(paste0())
  
  # Isolating markers for the given cluster
  temp <- results[[i]]
  
  # Only keeping significant markers
  temp <- temp[temp$p_val_adj < .05,]
  
  # Separating up & down markers
  up_markers <- rownames(temp[temp$avg_log2FC > 0,])
  down_markers <- rownames(temp[temp$avg_log2FC < 0,])
  
  # Printing a statement saying how many markers we have
  print(paste0("Cluster ", substr(i,2,nchar(i)), " contains ", dim(temp)[1], ' markers'))
  
  if(dim(temp)[1] == 0){
    next
  }
  
  print(paste0('Up in WT = ',
               length(up_markers)))
  
  print(paste0('Up in MUT = ',
               length(down_markers)))
  
  # Adding the markers/gene to the list indexed by cluster
  marker_list[[paste0(i,'_wt')]] <- up_markers
  marker_list[[paste0(i,'_mut')]] <- down_markers
  # break
}
```


# MSigDB 

We will use the molecular signatures database ([MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/)) to load their mouse collections to then run enrichment analysis with clusterProfiler

## Database generation

In this chunk of code we will generate the database for use in clusterProfiler.

```{r}
# Loading all gene sets present in MSigDB for mouse
all_gene_sets <- msigdbr(species = 'Mus musculus')

head(all_gene_sets) # looking at the first six rows of the data
```

Next we will look at the different collections in MSigDB.

```{r}
unique(all_gene_sets$gs_collection_name)[order(unique(all_gene_sets$gs_collection_name))]
```

More details about the collections are available.

```{r}
head(msigdbr_collections())
```


```{r}
# Collecting unique rows in the database and creating the term to gene object
msigdbr_t2g <- dplyr::distinct(all_gene_sets, gs_name, gene_symbol)
```


## Running enricher

Next we will run the enricher function on the DEGs between conditions. This step will take a while. 

```{r}
pathway_results <- list()
for (i in names(marker_list)){
  print(i)
  
  res <- enricher(gene = marker_list[[i]], # cluster marker genes
                  TERM2GENE = msigdbr_t2g, # reference database
                  universe = rownames(obj), # background reference list
                  pvalueCutoff = 1)
  
  pathway_results[[i]] <- res
}

```

## Looking at results for cluster 1

First we will look at the pathways enriched in WT.

### Enriched in WT

```{r}
temp <- pathway_results[['c1_wt']]

temp
```

```{r}
# Put it in dataframe form to query and plot
temp <- as.data.frame(temp) 

dim(temp)

head(temp)
```

Here we have the gene set name but not the collection they belong to, so we'll need to map back to that.

```{r}
# creating a dataframe with gene sets mapped to their collection name
collections <- as.data.frame(dplyr::distinct(all_gene_sets, gs_name, gs_collection_name))

# # Checking that they all overlap
# table(temp$ID %in% collections$gs_name)

rownames(collections) <- collections$gs_name
temp$collection <- collections[temp$ID,]$gs_collection_name

# Looking at how many entries we have for each collection in our results
table(temp$collection)

# Looking at the number of significant entries (TRUE) for each collection
table(temp$collection,temp$p.adjust < .05)
```

#### WikiPathways Results

Next we'll view and plot the WikiPathways results, one of the collections in MSigDB.

```{r}
wiki <- temp[temp$collection == 'WikiPathways',] # subsetting to just Wiki results

wiki$id <- gsub('WP_','', wiki$ID) # removing the WP_ preceder from term IDs
wiki$id <- substr(wiki$id,1,30) # limiting to thirty characters for ID bc some get long
wiki$id <- factor(wiki$id, levels = rev(unique(wiki$id))) # ordering them by p-vals

ggplot(wiki[1:10,], # only plotting the first ten 
       aes(x = -log10(p.adjust), 
           y = id,
           color = FoldEnrichment,
           size = Count)) +
  theme_bw() + # my theme of choice
  geom_point() +
  scale_color_gradient(low = 'pink', high = 'darkred') + # creating a monotonic color scheme
  # creating a significance cutoff to let viewers know all values plotted are significant
  geom_vline(xintercept = -log10(0.05), colour = 'lightgrey', linetype = 2, linewidth = 1) +
  # increasing the xlims because the larger points were getting cut off
  xlim(c(-log10(0.05),-log10(min(wiki$p.adjust)) + 1)) +
  # cleaning up the labels on the graph
  labs(fill = 'Fold\nEnrichment', 
       size = 'Associated\nGenes') +
  xlab('-log10 FDR') +
  ylab("WikiPathway Terms") +
  ggtitle("Top Ten WikiPathway Terms Associated with C1 WT") 

wt_c1 <- wiki[1:10,] # saving to plot with enriched in MUT
```

### Enriched in MUT

```{r}
temp <- pathway_results[['c1_mut']]

temp
```

```{r}
# Put it in dataframe form to query and plot
temp <- as.data.frame(temp) 

dim(temp)

head(temp)
```

Here we have the gene set name but not the collection they belong to, so we'll need to map back to that.

```{r}
# creating a dataframe with gene sets mapped to their collection name
collections <- as.data.frame(dplyr::distinct(all_gene_sets, gs_name, gs_collection_name))

# # Checking that they all overlap
# table(temp$ID %in% collections$gs_name)

rownames(collections) <- collections$gs_name
temp$collection <- collections[temp$ID,]$gs_collection_name

# Looking at how many entries we have for each collection in our results
table(temp$collection)

# Looking at the number of significant entries (TRUE) for each collection
table(temp$collection,temp$p.adjust < .05)
```

#### WikiPathways Results

Next we'll view and plot the WikiPathways results, one of the collections in MSigDB.

```{r}
wiki <- temp[temp$collection == 'WikiPathways',] # subsetting to just Wiki results

wiki$id <- gsub('WP_','', wiki$ID) # removing the WP_ preceder from term IDs
wiki$id <- substr(wiki$id,1,30) # limiting to thirty characters for ID bc some get long
wiki$id <- factor(wiki$id, levels = rev(unique(wiki$id))) # ordering them by p-vals

ggplot(wiki[1:10,], # only plotting the first ten 
       aes(x = -log10(p.adjust), 
           y = id,
           color = FoldEnrichment,
           size = Count)) +
  theme_bw() + # my theme of choice
  geom_point() +
  scale_color_gradient(low = 'pink', high = 'darkred') + # creating a monotonic color scheme
  # creating a significance cutoff to let viewers know all values plotted are significant
  geom_vline(xintercept = -log10(0.05), colour = 'lightgrey', linetype = 2, linewidth = 1) +
  # increasing the xlims because the larger points were getting cut off
  xlim(c(-log10(0.05),-log10(min(wiki$p.adjust)) + 1)) +
  # cleaning up the labels on the graph
  labs(fill = 'Fold\nEnrichment', 
       size = 'Associated\nGenes') +
  xlab('-log10 FDR') +
  ylab("WikiPathway Terms") +
  ggtitle("Top Ten WikiPathway Terms Associated with C1 MUT") 

mut_c1 <- wiki[1:10,] # saving to plot with enriched in MUT
```

### Plotting Both

```{r}
wt_c1$group <- 'WT'
mut_c1$group <- 'MUT'

temp <- rbind(wt_c1,
              mut_c1[10:1,])

temp$id <- factor(temp$ID, levels = unique(temp$ID))

ggplot(temp) + # only plotting the first ten 
  facet_grid(rows = vars(group),
             scales = 'free', space = 'free') +
  theme_bw() + # my theme of choice
  geom_point(aes(x = -log10(p.adjust), 
           y = id,
           color = FoldEnrichment,
           size = Count)) +
  scale_color_gradient(low = 'pink', high = 'darkred') + # creating a monotonic color scheme
  # creating a significance cutoff to let viewers know all values plotted are significant
  geom_vline(xintercept = -log10(0.05), colour = 'lightgrey', linetype = 2, linewidth = 1) +
  # increasing the xlims because the larger points were getting cut off
  xlim(c(-log10(0.05),-log10(min(wiki$p.adjust)) + 1)) +
  # cleaning up the labels on the graph
  labs(fill = 'Fold\nEnrichment', 
       size = 'Associated\nGenes') +
  xlab('-log10 FDR') +
  ylab("WikiPathway Terms") +
  ggtitle("Top Ten WikiPathway Terms Associated with C1") 
```

