---
title: "Pathway Analysis w/ MSigDB"
author: "D. Ford Hannum"
date: "`r Sys.Date()`"
output: 
        html_document: 
          toc: yes
          toc_float: yes
          theme: united
          highlight: tango
          # code_folding: hide
          number_sections: true
          keep_md: no
          toc_depth: 5
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# Set the directory to where everything is located
knitr::opts_knit$set(root.dir = '/Users/dfhannum/Desktop/OphthalmApp/') # sets it for all code chunks

library(Seurat) # for scRNA functions and plots
library(ggplot2) # for custom plotting
library(data.table) # data file processing and tstrsplit
library(clusterProfiler) # pathway analysis

library(org.Mm.eg.db) # <org.Hs.eg.db> human or <org.Rn.eg.db> rat

library(msigdbr) # access to MSigDB
```

# Introduction

In this document we are going to use an already created Seurat scRNA object. We are then going to identify cluster markers, and submit those cluster markers for over-representation (enrichment) analysis with MSigDB. We will then make some plots representing the most significant pathways.


# Data

I'm using a previously published dataset that I've reanalyzed for other example scripts.

```{r}
# Loading in the data
obj <- readRDS('./data/seurat_obj_full.Rds') 
```

```{r}
# Visualizing the data
DimPlot(obj)
```

We are going to rerun the clustering to identify a smaller number of clusters. Many of these clusters will share similar transcriptomic profiles. For the speed/visualizations of my downstream example it'll be easier working with fewer clusters.

```{r}
obj <- FindClusters(obj, resolution = .05)
```

```{r}
# Projecting the data in umap space
DimPlot(obj)
```

# Finding Cluster Marker Genes


```{r}
# Useful function to find markers for each cluster compared to all others
markers <- FindAllMarkers(object = obj,
                          only.pos = T, # focusing on upregulated markers
                          test.use = 'wilcox', # the default test that runs quickly
                          logfc.threshold = .2 # setting a strong cutoff 
                          )
```

Next we are going to extract the differentially expressed genes for each cluster

```{r}
# Creating an empty list to add to
marker_list <- list()

# A for loop to loop through the clusters
# You can replace obj$seurat_clusters with your cluster identity
for (i in levels(obj$seurat_clusters)){
  
  # Print statements are always useful in for loops to track progress and to 
  # make sure it's doing what it's suppose to
  print(paste0("Cluster ", i))
  
  # Isolating markers for the given cluster
  temp <- markers[markers$cluster == i,]
  
  # Only keeping significant markers
  temp <- temp[temp$p_val_adj < .05,]
  
  # Printing a statement saying how many markers we have
  print(paste0("Contains ", dim(temp)[1], ' markers'))
  
  # Adding the markers/gene to the list indexed by cluster
  marker_list[[i]] <- temp$gene
}
```


# MSigDB 

We will use the molecular signatures database ([MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/)) to load their mouse collections to then run enrichment analysis with clusterProfiler

## Database generation

In this chunk of code we will generate the database for use in clusterProfiler.

```{r}
# Loading all gene sets present in MSigDB for mouse
all_gene_sets <- msigdbr(species = 'Mus musculus')

head(all_gene_sets) # looking at the first six rows of the data
```

Next we will look at the different collections in MSigDB.

```{r}
unique(all_gene_sets$gs_collection_name)[order(unique(all_gene_sets$gs_collection_name))]
```

More details about the collections are available.

```{r}
head(msigdbr_collections())
```


```{r}
# Collecting unique rows in the database and creating the term to gene object
msigdbr_t2g <- dplyr::distinct(all_gene_sets, gs_name, gene_symbol)
```


## Running enricher

Next we will run the enricher function on the top markers from each cluster.

```{r}
pathway_results <- list()
for (i in names(marker_list)){
  print(i)
  
  res <- enricher(gene = marker_list[[i]], # cluster marker genes
                  TERM2GENE = msigdbr_t2g, # reference database
                  pvalueCutoff = 1)
  
  pathway_results[[i]] <- res
}

```

## Looking at results for cluster 1

```{r}
temp <- pathway_results[[1]]

temp
```

```{r}
# Put it in dataframe form to query and plot
temp <- as.data.frame(temp) 

dim(temp)

head(temp)
```

Here we have the gene set name but not the collection they belong to, so we'll need to map back to that.

```{r}
# creating a dataframe with gene sets mapped to their collection name
collections <- as.data.frame(dplyr::distinct(all_gene_sets, gs_name, gs_collection_name))

# # Checking that they all overlap
# table(temp$ID %in% collections$gs_name)

rownames(collections) <- collections$gs_name
temp$collection <- collections[temp$ID,]$gs_collection_name

# Looking at how many entries we have for each collection in our results
table(temp$collection)

# Looking at the number of significant entries (TRUE) for each collection
table(temp$collection,temp$p.adjust < .05)
```

### WikiPathways Results

Next we'll view and plot the WikiPathways results, one of the collections in MSigDB.

```{r}
wiki <- temp[temp$collection == 'WikiPathways',]

wiki$id <- gsub('WP_','', wiki$ID)
wiki$id <- factor(wiki$id, levels = rev(unique(wiki$id)))
ggplot(wiki[1:10,],
       aes(x = -log10(p.adjust),
           y = id,
           color = FoldEnrichment,
           size = Count)) +
  theme_bw() +
  geom_point() +
  scale_color_gradient(low = 'pink', high = 'darkred') +
  geom_vline(xintercept = -log10(0.05), colour = 'lightgrey', linetype = 2, linewidth = 1) +
  xlim(c(-log10(0.05),-log10(min(wiki$p.adjust)) + 1)) +
  labs(fill = 'Fold\nEnrichment', 
       size = 'Associated\nGenes') +
  xlab('-log10 FDR') +
  ylab("WikiPathway Terms") +
  ggtitle("Top Ten WikiPathway Terms Associated with Cluster 1") 
```

